---
title: "Grade 6 Maths"
author: "Prem"
date: "Monday, April 11, 2016"
output:
  html_document:
    highlight: pygments
    theme: spacelab
    toc: yes
---

```{r fig.height = 3, fig.width = 4, fig.align='center', echo=FALSE, warning=F}
# install grid and png packages if not present
library(png)
library(grid)
```

https://www.khanacademy.org/math/in-sixth-grade-math
https://www.khanacademy.org/profile/peterwcollingridge/

http://betterexplained.com/articles/learning-how-to-count-avoiding-the-fencepost-problem/

# Chapter 1: Knowing our numbers

## Comparing numbers

* **Writing Numbers in Expanded form**
    * write $14,987$ in expanded
    * $1 * 10,000 = 10,000$
    * $4 * 1,000 = 4,000$
    * $9 * 100 = 900$
    * $8 * 10 = 80$
    * $7 * 1 = 7$
    * $10000+4000+900+80+7 = 14,987$

## Large Numbers in Practice

* In this topic we'll look at the basics of **number approximation**, when we round numbers to be less exact 
* This would happen when you don't trust your measurements or want to simplify things, in other words when you want to **roughly** know the number 

### Rounding to nearest 10
    
* What is the nearest multiple of 10 for a given number? 
* Look at value in **ones column**, one place to the right of the **tens**, and decide which multiple of ten is closest, then Round up or Round down
* multiples of ten = $10,20,30,40,50,60,70,80,90$
* `value:nearest multiple of ten` = 36:40, 34:30, 65:70, 92:90, 12:10
```{r fig.height = 3, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("grade6_fig/C1_F1.PNG"))
```
* ***note***: Rule of Rounding 5: When rounding number 5, which is 5 away from one multiple and 5 away from the other, the rule is to round up
    * This is just a rule
    
### Rounding to nearest 100

* Look at value one place to the right of the **hundreds** column, in the **tens** column and decide which multiple of hundred is closest.
* `value:nearest multiple of ten/hundred` = $154:150/200,\\  4,674:4,670/4,700,\\  9,995:10,000/10,000,\\  8,346:8,350/8300$
```{r fig.height = 3, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("grade6_fig/C1_F2.PNG"))
```

### 2 step estimation problem

* The Pokemon day-care centre has 128 orphaned Pokemon. They've just been given another 54. The daycare wants to share these Pokemon amongst the 43 people who have signed up to take a Pokemon. **Roughly speaking** how many Pokemon will each trainer get?  
    1. Round 128 and 54 to the nearest ten
        * 120 + 50 = 170 Pokemon
    2. Round the amount of trainers to the nearest ten
        * 40
    3. Divide the amount of Pokemon by amount of trainers 
        * Let `P` be amount of Pokemon
        * 170 / 40 = P 
        * P x 40 = 170. Solve for P
        * List Multiples of 40: 40, 80, 120, **160**, 200.
        * 170/40 = 4 Pokemon. Each Trainer will receive roughly 4 Pokemon.

### Multiplication estimation example

* A ticket agent sells 42 tickets. Tickets cost £28 each. Use rounding to estimate the total amount of money from sales
    1. Round number of tickets and cost of tickets to nearest ten
        * 40 tickets
        * £30 each
    2. Multiply together
        * 40*30 = 4*3 with two zeroes added on the end (12 -> 1,200)
        * Roughly we will have £1,200 from sales 
        
## Using Brackets

### Constructing Numerical Expressions with Parathesis 

* Mike and Amon are having a beyblade battle. Amon has 3 Beyblades and wins 1 more. He then battles Mike for all or nothing, doubling his amount of Beyblades. **Write an expression to model this situation without performing any operations**
    * $2(3+1)$
    
# Chapter 2: Whole Numbers

##  Properties of Whole Numbers

### Commutative Law of addition

* **commutative law** = order doesn't matter if you're adding a bunch of things. i.e. you can add numbers in any order and the result will always be the same
    * $10+20+4=34$
    * $4+10+20=34$
    
### Commutative law of multiplcations

* **commutative law** = numbers can be multiplied in any order and the result will always be the same

### Associative law of addition

* **associative law of addition** =  It doesn't matter which numbers are added first in paranthesis, the result will always be the same
    * i.e. it doesn't matter if you associate the 20 with the 1, or the 1 with the 4. 
    * (20+1)+4 = 25
    * (1+4)+20 = 25
    
### Associative law of multiplication

* **Associative law of multiplication** = It doesn't matter which numbers are added first in brackets, the result will always be the same

### Distributive property over addition

* **distributive law of multiplication over addition** = distributive the number you are multiplying by to each number in the brackets
    * 4(8+3). Distribute the 4 to each number
    * (4*8)+(4*3) 
    * 32+12 = 44
* **Why use distributive law instead of just BODMAS (adding numbers in brackets first)?**
    * Using the distributive law helps you in algebra, when you will have variables 
    * **example 1**:
    * 4(8+x)=44. Solve for `x`
    * (4*8) + (4*x) = 44
    * 32 + 4x = 44
    * 4x = 12
    * x = 3
    * **example 2**:
    * Each month you pay the following bills: Water:$30, Electricity:$150, Cable:$70. How much do you pay in a year and how much is each item?
    * 12(30w,150e,70c)
    * 360w + 1800e + 840c = $3000.
    * We could of added the bills together, then multiplied [12(250) = 3000], but you would of lost crucial information about how much **each** bill costs
    * **Related Branches**: Factoring, Polynomials, Algebra. Knowing the distributive law will be useful when you reach those branches
    
### Distrbutive property over substraction

* **distributive law of multiplication over substraction** = distributive the number you are multiplying by to each number in the brackets
    * $5(9-4)$
    * $5*9-5*4$
    * $45-20$
    * $25$

### Summary

* **Commumative Law** = a+b+c = b+c+a
* **Associative law** = a+(b+c) = b+(a+c)
* **Distributive property** = a+b(c+d) = a+bc+bd
    

# Chapter 3: Playing with Numbers

* In this chapter we will learn about factors and multiples and about the divisibility rules for 2, 3, 4, 5, 6, 8, 9 and 10

## Factors and Multiples

* In this topic we will see that a factor of a number is an exact divisor of that number, and a multiple is a result of the multiplication of that number 

### Finding Factors of a Number

* In this tutorial we'll begin to look at the numbers that "make up" the number
* **Use**: This will be useful throughout maths, whether we are adding up fractions, exploring mystical numbers patterns or breaking computer codes, factoring numbers are key!

* **factor** = the factory that *makes* a whole number from parts
    * ***note***: factor comes from latin and makes do/make. The factor makes something happen. In math, a factor makes a result happen, it produces the result
    * i.e. 2 and 6 make 12 happen / 2 and 6 produce a result of 12 (2*6= 12)
    
* **Finding Factors**
    1. Write the smallest and largest factor
    2. Work up the number line(1:10), filling in the gaps between the smallest and largest factor
    3. Keep going until the gaps are filled

* **Recognizing Divisibility**: is x divisible by...
    * Number 1. All Whole Numbers are divisble by 1
    * Number 2. The number is even (i.e, the last digit is 2, 4, 6, 8, 0)
    * Number 3. The digits add up to a multiple of 3.
        * If you're unsure if the sum is multiple of 3, you can check the sum it self by adding the digits together. 
        * I.e. Is 386, 802 divisble by 3? 3+8+6+8+0+2=27. Not sure if 27 is multiple of 3? 2+7=9. Yes, 9 (therefore 27) is multiple of 3.
    * Number 4. The last two digits are a multiple of 4
    * Number 5. The last digit is zero or 5
    * Number 6. Divisible by both 2 and 3
    * Number 7. 
    * Number 8. The last three digits are divisble by 8
    * Number 9. The digits add up to a multiple of 9
    * Number 10. The last digit is zero

* **Find All Factors of 120** 
   * $120 = 1 * 120$
   * $120 = 2 * 60$. Is x divisible by 2? Check if one's place is even.
   * $120 = 3 * 40$. Is x divisible by 3? Add up its digits and see if sum is divisible by three. I.e. 1+2+0 = 3. Yes, 120 is divisible by 3. 
   * $120 = 4 * 30$ Is x divisible by 4? You ignore everything beyond hundreth's place and look at last two digits.i.e., is 20 divisible by 4, so 120 is also divisible by 4.
   * $120 = 5 * 24$ Is x divisible by 5? If last digit is 0 or 5, then yes
   * $120 = 6 * 20$. Is x divisible by 6? If divisible by 2 and 3, yes
   * $120 = 8 * 15$ Is x divisible by 8? If last 3 digits divisble by 8, yes
   * $120 = 10 * 12$ Is x divisble by 10? If last digit is 0, then yes
   * Factors of 120 are: 1   2   3   4   5   6   8  10  12  15  20  24  30  40  60 120

```
FUN <- function(x) {
    # Convert number to integer to speed up function
    x <- as.integer(x)
    # Find all numbers to divide by (1:x)
    div <- seq_len(abs(x))
    # Only return numbers where division results in no remainder
    # 0L, the `L` notation ensures number is stored a a integer
    # not a double
    factors <- div[x %% div == 0L]
    # Make a list with negative and positive factors
    factors <- list(neg = -factors, pos = factors)
    return(factors)
}
```

### Find Multiples Of A Number

* To find a multiple the best way is just simply list the multiples by doing your times table
* factor x factor = multiple

## Prime and Composite Numbers

* In this topic we will see that composite numbers have more than one factor and prime numbers have just two: 1 and the number itself
* Prime numbers have been studied by mathematicians and mystics for eons. By studying them you will unfold more fascinating views of the universe
* **use**: Prime numbers are used in cryptography and studying nature, it's what decides how many petals a flower has and what stops theives from entering your online bank account

### Prime Numbers

* Prime numbers are the building blocks of numbers, that can not be broken down into products of smaller numbers
* **prime number** = A natural number (1,2,3...) divisible by **exactly two** natural numbers, 1 and itself
    * 1 = not prime. Only divisible by one number, not exactly two
    * 2 = prime number. Is divisble by 1 and it self
        * 2 is the only even number that is prime
    * 
* **Identifying Prime Numbers**
    * Divide number by all the numbers between 1 and it self. If the number is only divisible by 1 and it self then it is prime
    
```
FUN <- function(x) {
    # Convert number to integer to speed up function
    x <- as.integer(x)
    # Find all numbers to divide by (1:x)
    div <- seq_len(abs(x))
    # Only return numbers where division results in no remainder
    # 0L, the `L` notation ensures number is stored a a integer
    # not a double
    factors <- div[x %% div == 0L]
    # Check If amount of factors is larger than 2
    if (length(factors) > 2) {
      print(paste(x,"is not prime"))
    } else {
    return(factors)
    }
}

# Which of these numbers is prime?

sapply(c(35,47,55,60,87), FUN)
```

## Test for Divisbillity of numbers

### The why of the three divisibillity

* Take Number, such as 498
* Re-write number as something that is "1 + something divisble by 3""
    * 498 -> 4(1+99) + 9(1+9) + 8
    * Instead of writing 4(100)...the trick is write 4(1+something-divisible-by-3) instead of 100
* Distribute the four and re-arrange terms
    * 4+(4*99)+9+(9*9)+8.
    * 4(99) + 9(9) + 4 + 9 + 8. # Re-arrange the terms
* Check if terms are divisible by 3
    * 4(99), divisible by 3 **because anything multipled by something divisible by three** is divisible by three!
    * 9(99), divisible by 3, because anything multiplied by something divisible ny three is divisble by three!
    * 4(99) + 9(99), If you add**two things which are divisible by three added, the whole thing is still going to by divisble by three**
    * 4 + 9 + 8 = 21. 2+1. **What are we left with? Our original digits!** Just need to make sure original digits are divisble by three
    * In other words, 498 is divisble by 3 if 4 + 9 + 8 is divisible by 3!

### Why can't I add all the digits to see if its divisble by four?

 read comments
 
 Say you have the number 19. That is 1 ten and 9 units. If you increase the value by one, you would have 10 in the units place. Since you can't have that you increase the tens place to 1 and decrease the units place by 10 to get 2 ten and 0 units. The sum of the digits therefore decreases from 11 (1 ten and 10 units) to 2 (2 ten and 0 units), which is a decrease of 9.


 https://www.khanacademy.org/math/in-sixth-grade-math/playing-numbers/test-divisibility-numbers/v/the-why-of-the-3-divisibility-rule

### The Why of the 9 divisbillity rule?

* is 2943 divisble by 9?

1. Break the number up by place value 
    * 2(1000) + 9(100) + 4(10) + 3
2. Re-write as "1 + something divisble by 9"
    * 2(1+999)+9(1+99)+4(1+9)+3
3. Distribute
    * 2+(2*999)+9+(9*99)+4+(9*4)+3
4. Re-arrange the terms
    *(2*999)+(9*99)+(9*4)+2+9+4+3
5. Check divisibillity of all parts
    * (2*999)+(9*99)+(9*4) = yes. Anything times something divisible by 9 is divisble by 9
    * 2+9+4+3 = yes. Second part needs to be divisble by 9 in order for the whole number (2943) to be divisible by 9. 
    * In other words, 2943 is divisble by 9 if 2+9+4+3 is divisible by 9!
### The why of the 4 divisibillity rule

* is 419108 disivible by 4? 
* A number is disible by 4 if the last two digits are divisble by 4

1. We-can re-write the number as a multiple of 100 plus the last two digits

    * 419108 = 4191 + 08 Because 419100 is a multiple of 100, it is also a multiple of 4
2. So long as the last two-digits are divisble by four then the original number must be divisble b four!

## Common Divisibillity

* Concepts = prime factors, division
* All numbers divisible by A and B are also divisible by...
* **example** All numbers divisble by 21 and 20 are also divisble by...
    * 333?
    * 30?
    * 172?
    * 147?
    * 222?
    1. Find Prime Factors for 21 and 20
    * 21 = 7 * 3. 20 = 5*2*2
    * Any number divisble by both 21 and 20 must have 2*2*3*5*7 as part of its prime factorization 
    2. Check if the prime factors of the options match
    * The prime factors of 30 is 2*3*5
    * Since 2*3*5 is a part of 2*2*3*5*7, all numbers divisble by 21 and 20 must be divisible by 30. 
    
## Prime Factorization 

* You know what prime numbers are and how to identify them, in this tutorial we will look at how all positive numbers can be broken down into products of prime numbers (In some ways, prime numbers as the atoms of the number world that can be multiplied to create any number!). Besides being a fascinating idea, it's extremely useful! Prime factorization can be used to decrypt encrypted information. 

* Concepts = division, prime factors. 

### Finding the prime factors

Find the prime factorisation of 75. Write your answer using exponential notation.

1. Work through lists of primes, finding the smallest prime number that will go into 75.
* **note**: this is where learning about divisibillity rules comes in handy
    * list of primes = 2,3,5,7,11...
    * Is 75 divisible by 2? No. Ones place is not even.
    * is 75 divisible by 3? Yes. 5+7=12, which is divisble by 3.
    * 75 = 3 * 25
    * Find the factors of 25.
    * is 25 divisble by 2? No. If it's not divisble by 75, then won't be by 25. 
    * is 25 divisble by 3? No. 2+5 = 3.
    * is 25 divisble by 5? Yes.
    * 25 = 5 * 5
    * We're done, we're left with only primes!
    * 75 = 3 * 5 * 5
2. Re-write using exponential notation
    * If we have repeated primes we can write those as exponents
    * 5*5 is the same thing as 5^2
    * 75 = 3 * 5^2
  
## Lowest Common Multiple 

* Life is good but it can be better. Just imagine being able to find the smallest number that is a multiple of two other numbers! Other than making your life more fufilling, it will allow you to do incredible things like adding fractions

* `concepts` = Prime Factors, Fractions, Addition

* `use` =
    * Adding fractions
    * Wastage. For example, if buns come in packs of ten and burgers in packs of 8. If you don't want to waste any you will need to buy the LCM of buns and burgers. e.g. LCM(10,8). This also occurs in manufacturing, where one company will only give you parts in packages of X and another will only give you parts in packes of Y. If you need one of each to make your product you will need to find the LCM(X,Y) to avoid waste parts.
    * Finding best deals.
  
* **Method 1: Prime Factorization**
    * What is the LCM of A and B?
    * Take Prime Factorization of A and B and construct smallest number whose ingredients is all the factors of both A and B, which will be the LCM
    * What is the LCM of 12 and 18?
    1. Find prime Factors of 12 and 18 using factorization trees
        * 12 = 2*2*3
        * 18 = 2*3*3
    2. Multiply of the factors that appear in both numbers and construct a new number
    * The LCM has a super-set containing all of these factors or all of these factors in it as many times as we have it in any one of these (A or B)
    * in order to be divisible by 12 you need to have two 2's and atleast one 3. But to be disivisible by 18 you need another 3! 2*3*3*2 is divisible by both 12 and 18. 
    * LCM(18,12) = 2*3*3*2 = 36
* **Method 2: Brute Force**
    * Go through times table for both numbers until you find LCM
* Method 1 (Prime Factorization) is better as firstly, you deconstruct the number (so it's fun!) and secondly, things will get hairer with larger numbers so it's easier, quicker and cleaner to have a systematic method
  
```{r fig.height = 3, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("grade6_fig/C3_F1.PNG"))
```

## Greatest Common Factor 

* You know how to find factors of a number. What about factors that are common to two numbers? Even better, imagine the largest factors that are common to two numbers. 

* **note**: It's better to just use method 2. 

* What is the GCF of A and B? 
    * Finding the GCF/GCD(greatest common *divisor*) is finding the largest number that divides into both numbers
* **method 1: Common Factors**:
    * List all factors for each number, and find the greatest common factor
    * What is the GCF of 10 and 7?
        * 10 = 1,2,5,10
        * 7 = 1,7
        * GFC(10,7) = 1
    * **note**: two numbers with only 1 as their GCF is known as **relatively prime**
* **Method 2: Common Prime Factors**:
    * List all prime factors for each number, and find the largest set of common prime factors
    * The greatest common factor will be a product of the set
    * What is the GCF of 30 and 15? 
        * 30 = 2,3,5
        * 21 = 7,3
        * GFC(30,31) = 3
    * What is the GCF of 105 and 30?
        1. Factor each number completely as a product of its primes 
            * 105 = 3, 5, 7
            * 30 = 2,3,5
        2. Find the common prime factors
            * Largest Set of Common Prime Factors = 3,5
        * GCF(105,30) = 3*5 = 15
        * 15 is the largest number divisible into both numbers
    * **note**: If numbers do not share any prime factors, then the GCF is 1 as this factor is common to all numbers
    * What is the GCF of 60 and 60?
        1. Factor each number completely as a product of its primes 
            * 160 = 2,2,3,5
            * 60 = 2,2,3,5
        * Both numbers share all of their factors, so GCF = 2*2*3*5 = 60
    * **note**: In general, the greatest common factor of any number and its self is that number 
    
## GCF and LCM word problems

* **Example 1**: There are 32 forwards and 80 guards in Leo's basketball league.
Leo must include all players on a team and wants each team to have the same number of forwards and the same number of guards.
If Leo creates the greatest number of teams possible, how many guards will be on each team?
    * In order to find out how many teams Leo can create, we need a number that is a factor of 32 and 80, so that the 32 forwards and 80 guards can be divided up evenly. 
    * So, if there were 4 teams, there would be 32/4= 8 forwards and 80/4= 20 gaurds on each team. This creates equal teams (the same number of players on each team) but it isn't the **greatest number of teams possible!**
    * To find the greatest number of teams, we need to find the GCF of 32 and 80. 
        * gfc(32,80) = 16
    * The greatest number of teams Leo can make is 16. 
    * Each team would have 32 / 16 = 2 forwards and 80/16 = 5 gaurds
    
* **example 2**: Professor Oak has 32 Pokeballs and 8 Rare Candies. He wants to give all of these in identical gift bags to new trainers. What is the most amount of gift bags he can make?

    * To find the greatest number of gift bags we need to find the Greatest Common Factor of 32 and 8 
    * In other words, find the GCF(32,8) = 8. 
    * There will be 8 identical packages, which contain 32/8 = 4 Pokeballs and 8/8 = 1 Rare Candy. 
    
* **example 3**: Ged milks the goats every 8 days and goes to market every 6. Today he did both on the same day. How many days until he milks and the goat and goes to the market on the same day? 
    * Let's look at Ged's diary, by going up the times table for each number. 
        * 8 = 8,16,24
        * 6 = 6,12,16,24
    * As we keep going we find that the multiples meet on the 24th day.
        * Mathematically we say that 24 is the Least Common Multiple of 8 and 16. LCM(8,16) = 24
    * It will be 24 days until ged milks the goats and goes to market on the same day
    * **note**: although I've listed the multiples of each number, which is the brute force method, for clarity, it is easier to simply work out the LCM via the Prime Factorization Method.

# Uses

## Factors 

### GCF

* Creating Identical Packages with two items(A and B)
    * Find GCF, divide each item by GCF
    * GCF = the greatest number of packages that can be made
    * item / GCF = amount of items in each package
        * A / GCF = number of item A in each package
        * B / GCF = number of item B in each package
* Splitting items A and B into identical rows
    * Find GCF, divide each item by GCF
    * GCF = the greatest number of rows that can be planted
    * item / GCF = amount of items in each row
        * A / GCF = number of item A in each package
        * B / GCF = number of item B in each package
### LCM

* **LCM** is great for finding where two events will occur together, by finding where the multiples of each number first meet 
* Preventing Wastage. item A and B is needed to build a product/make a hotdog. You can only buy item A in packs of X and item B in packs of Y.
    * Find LCM, divide LCM by each pack
    * LCM(X,Y) = smallest total number of item A/B you can buy without wasting. 
        * I.e. the amount of items you need of A and B each to not have any left-overs. 
    * LCM / packs = amount of packs you need to buy of each item
        * LCM / X = amount of packs you need to buy of item A
        * LCM / Y = amount of packs you need to buy of item B
* Finding over-lapping events
* Finding minimum over-lapping scores. 
    * Pikachu gains 8 points with every match. Squirtle gains 9 points. At the end of the day, each pokemon has the same amount of points. 
    * Find LCM.
    * LCM = Least number of total points that person A and B could of each achieved 
    * Pikachu gains 8 points with every match. Squirtle gains 9 points. At the end of the day, each pokemon has the same amount of points. What is the least amount of matches Pikachu could of had? 
    * Find LCM, divide LCM by points per event for each person
    * LCM = Least number of total points that person A and B could of each achieved 
    * LCM / points per event = amount of events that each person did in order to reach the same score 
    
# Chapter 4: Basic Geometrical Ideas

## Introduction

* **Euclid Father Of Geometery** = Euclid, born 2,500 years ago in Alexandira was the first person to prove something beyond doubt. Not simply have a good feeling but prove that something is wholly correct for the whole universe.   

While he wasn't the first to study geometery, many peoples would of before him simply by looking at the shapes around us in nature and considering what shapes will help us build structures, like pots or even Pyramids. However in his book he essentially under-pinned modern maths which is begining with an basic assumption, known as a theorem, and then going to prove it, known as a proof.  Thus, Euclid's legacy lies in turning theorems, stuff you feel pretty sure about, into proofs, stuff you know beyond a shadow of a doubt! 

## Terms & Labels in Geometery

* **geometery** = *geo* is earth and *metery* is measure, so it is the measurement of earth/nature

* In geometery we have a few attributes(ideas/things) that apply to all geometric terms
    * dimension = the amount of options to travel in
    * label = to identify terms we use labels rather than colours, sounds or pictures.
* **point** = a single dot which is fixed in one place. 
    * dimension = 0. A point can't travel in any direction or it will cease to be that point, so it has zero dimensions
    * label = $A$. The label for a point is a single letter within the Alphabet, one point could be A, another B
* **line segment** = two dots connected by a line / a part of a line bounded by two end points. Line *segments* have an end-point so are **finite* in length
    * dimensions = 1. A line segment is able to move back and forth, i.e. left and right
        * note: the geometric line has no width and only length, although in reality strings and sticks do have a width.
    * label = $\overline{AB}$, if you have two points (A&B) connected by a line 
    * **end point** = the points at the end of a line segment (A&B)
    * Line Segments are the most common things in geometery, as you will study the sides of shapes and distance between point. Anything with a finite length is a line segment

* **ray** = a line segment which starts at one point and extends forever in one direction 
    * dimensions = 1. Back and forth long the line
    * label = $\overrightarrow{AB}$: a line that extends past point B. *note* that the order of points matters, e.g. $\overrightarrow{BA}$ means a line that extends past point A.
    * vertex = the starting point of a ray

* **line** = a line segment that extends forever in both directions. Lines have no start/end point. 
    * dimensions = 1
    * label =  $\overleftrightarrow{AB}$
    * **midpoint** = a point that is exactly half-way between two points. 
    * **collinear** = points that sit on the same line, pretty much all points and midpoints along a given segment. 
    
* **planar/planes** = a infinite 2D space. Extends in all directions forever
    * dimension = 2. It can travel along 2 ways, up/down and left/right.
    * **planar segment** = piece of a paper/computer monitor. It is 2D, but finite

### Measuring Segments

* **congruent** = two line sigments with the same length

## Parallel and perpendicular lines

* **parallel** = lines that intersect at a right angle (90 degrees). (Makes a X shape) 
* **perpendicular** = two lines that are on the same plane and neve intersect, like rail-roads. 
* **intersecting lines** = lines that intersect but not at 90 degree angles

### Identifying parallel and perpendicular lines

* **identifying parallel lines** = if two suspected parallel lines intersect a third line at the same angle, then they are parallel 

## Angles

Angles are formed where corners are made. 

* **angle** = when 2 rays share a common endpoint an angle is made. The vertex is the angles heart. 
* **labelling angles** = $\angle ABC$: A&C = endpoints, B = vertex/the angles' heart.
* **naming angles**
    * **acute** = less than 90 degrees
    * **obtuse** = more than 90 degrees
    * **right angle** = 2 rays perpendicular to each other
### Measuring Angles

* angle measurement = how open or closed an angle is
* **how to measure**: Angles can be measured in **degrees** or **radians**
    * Degrees comes from the convention of using circles, therefore one method of measuring angles is **circle arcs**, where a segment is taken of a circle.
* **the length of a circle is 360 degrees**
    * Why is it 360 degrees? No one knows. However, there are hints in history and the way the universe works (sun).
    * There are 365 days in a year. 360 is quite close and is divisible by more things, perhaps this is way many ancient astronomers and civilisations (persians/mayans) used 360 day calenders
    * The circle being 360 degrees is a *convention* history has handed us.

* **measuring angles in degrees with circle arcs**
    1. Draw a circle with a point in the middle
    2. Draw a angle (2 rays with a common end point/vertex)
        * Make sure the angle's vertex is placed at the circle's midpoint
        * In the diagram below, the white line is the first ray and the green line is the second ray
    3. The fraction of the circle's circumference that is intersected by the 2 rays is the angle's measurement.
        * The segment of the circle that the two rays/angle slices out is the angle's measurement
    
```{r fig.height = 3, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("grade6_fig/C3_F4.PNG"))
```


## Circles

### Circles Glossary

* Formal definition of a circle. Tangent and secant lines. Diameters and radii. major and minor arcs

* **circle** = set of all points a fixed distance from A
    * Draw a point named `A`. `A` is the centre point, draw points in every possible posistion that are 2cms from `A`. You now have a circle
    * **centre point** = the point in the *middle* of the circle that is a equal distance from every point *on* the circle itself
        * Point `A` in the diagram below
* **radius** = line between centre-point and circumference (any point on the circle itself)
    * $\overline{AB} = 2cm$
* **tanget** = line that only interescts (touches) one point *on* the circle
    * $\overleftrightarrow{CM}$
    * Line `L` is tanget to the circle centered at `A`
    * Important to state circle's centre-point or else it could be refer to any circle!
* **secent** = line interesects (touches) two points on the circle
    * $\overleftrightarrow{DF}$
* **cord** = line segment between two points, a finite secent
    * $\overline{DF}$
* **diameter** = cord that passes through circle's centre-point
    * diameter is composed of 2 radii 
    * $\overline{CG}$
    * `CG` = `CA` + `AG`
* **arc** = line between two points *along* the circle
    * **minor arc** = quickest path between two points
        * $\frown {BG}$
    * **major arc** = longer path between two points
        * $\frown {BFG} = \frown {BDG}$

```{r fig.height = 3, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("grade6_fig/C4_F1.PNG"))
```

# Chapter 5: Unerstanding elementary shapes

* corners, edges, planes, open curves and closed curves are different shapes we seen around us. Understanding these shapes can help us solve a wide variety of real life problems

## Angles: acute, obtuse and reflex

* **acute** = angle less than 90 degrees. Rays are close to each other
* **right-angle** = 90 degrees. Rays are perpendicular to each other
* **obtuse** = more than 90 degrees. 

## Measuring Angles

* To review, a circle is split into 360 equal parts. Each part is called a degree, which is one way we measure angles. 

### Decomposing Angles

* To find out the measure of an angle, you can add or substract other angles

```{r fig.height = 3, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("grade6_fig/C4_F2.PNG"))
```

### Classifying Triangles

* There are **two methods** to classify triangles based on either the number of **equal sides** or the **angle measure**

* **number of equal sides**
    * **scalene** = triangle with no equal sides
        * no sides are congruent
    * **isosceles** = triangle with at least two equal sides
        * two sides are congruent
    * **equilateral** = triangle with all equal sides
        * all sides are congruent 
        * Yes, an equilateral is does also count as a isosceles
    
* **angle measure**
    * **acute** = all angles under 90 degrees
    * **right angle** = one angle is 90 degrees
    * **obtuse** = one angle is over 90 degrees
    * ***note***: all angles add up to 180

* When you don't know the lengths of the sides the angles come in handy.
    * **scalene** = no angles are equal
    * **isoceles** = 2 angles are equal
    * **equilateral = all angles are equal (60 degrees)

```{r fig.height = 3, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("grade6_fig/C5_F1.PNG"))
```

## Quadrilaterals 

### Quadrilateral Glossary

* **quadrilateral** = any shape with four shapes
* **concave** = has interior angle larger than 180 degrees. This means none of the sides can be parallel to each other
    * **note**: if angles were equal to 180 degrees, rather than more (concave) or less (convex) then it would be straight lines creating a triangle 
* **convex** = all of the interior angle less than 180 degrees.
    * **trapezoid** = exactly one pair of parallel sides
    * **parallelogram** = two pairs of parallel sides
        * **rectangle** = all 4 angles are a right angles
        * **rhombus** = all 4 sides are equal
        * **square** = rhombus + rectangle
* **note**: sum of any quadrilateral is 360 degrees

```{r fig.height = 3, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("grade6_fig/C5_F2.PNG"))
```

# Chapter 6: Integers

## Integers

* If you put the whole numbers and the negative numbers together you get a new collection of numbers known as integers.

### Negative Integers

* **negative number** = a number that is less than zero
* There are two ways to conceptulize negative numbers. One is a lack of something, such as a lack of temperature (freezing) or a lack of money (debt), the other is distance in relation to zero. 

* **distance from zero** = there are two parts to a number. It's absolute value and it's symbol. 
    * The **absolute value** tells how far the number is from zero
    * The **symbol** tells you in which direction. `negative` = left of zero, `positive` = right of zero.

* **lack of something** = negative numbers help us model real life situations, such as having debt. If you owe someone $2 and have a balance of $1, then you have a balance of -$1. While -$1 doesn't actually exist, it does help us keep track of of things, such as owing money or loosing health in a game. 

* ***note***: there is a difference between the **operation** of subtraction and the **object**(a negative number), even though both use the same sign. You should therefore say *subtract negative 3* and not *minus minus 3* to keep things clear. 

* **[history](https://nrich.maths.org/5961)**: 
    * rules for dealing with negative numbers first appeared in 7th century India, written down by Brahmagupta. 
    * However, references to negative numbers existed during 200 BCE in China, where positive numbers were items sold (money in) and negative were items brought (money out)
    * Greeks did not address negative numbers, as they dealt with geometery. Lengths, areas and volumes of shapes all had to be positive. Futhermore, their ideas were based on magnitudes. Magnitudes were represented by a line or an area, not by a number. 
    * Negative numbers appeared in Europe by the 15th century, after islamic and byzantine sources were translated
    
    ```
    A debt minus zero is a debt.
        a negative substract zero is a negative
    A fortune minus zero is a fortune.
        a positive substract zero is a positive
    Zero minus zero is a zero.
        zero subtract zero is zero
    A debt subtracted from zero is a fortune.
        a zero subtract a negative is a positive
    A fortune subtracted from zero is a debt. 0-(pos) = neg
        a zero subtract a positive is a negative. 0-(neg) = pos
    The product of zero multiplied by a debt or fortune is zero.
        zero multiplied by a negative or positive is zero
    The product of zero multiplied by zero is zero.
        zero times zero is zero
    The product or quotient of two fortunes is one fortune.
        positive * positive = single positive 
    The product or quotient of two debts is one fortune.
        negative * negative = single positive 
    The product or quotient of a debt and a fortune is a debt.
        negative * positive = negative
    The product or quotient of a fortune and a debt is a debt.
    ```
## Addition of Integers with different signs

* There are two parts to a addition/subtraction sum. The symbol which represents the **object**(a positive or negative number) and the symbol which represents the operation(subtraction or addition)

* -2+5 = 3
    * Method 1: Go 2 to the left of zero, then 5 to the right 
    * Method 2: What is the difference between 2 and 5? 
    
## Subtraction of integers

### Subtracting a negative number = adding a positive number

* Steve is in a debt of $3
    * steve: -3
* Steve's Uncle feels bad and wants him to have a balance of zero, so he takes away the negative 3 
    * -3 -(-3) = 0 
* Another way of saying this is that he gives him $3 dollars
    * -3 + 3 = 0. 
* Both are equal and mean the same thing. **subtracting a debt is the same as adding a fortune**

* **subtraction** = adding a negative number
    * 5-4 = 5+(-4)
        * -4 + 5 is also commutative, all 3 equations mean the same thing! 
* **addition** = adding a positive number 
    * 5+4 = 5+(+4)
    
* **subtracting a negative number** = taking away a debt / adding a positive
    * 3-(-2) = 5
    * Can also think of it as the negative signs cancel out
    
### Negative number word problem: Temperature 

*  The coldest temperature recorded in england was  -27 °C. The warmest temperature 37.4C. What is the difference in temperature?
    * **To find the difference between two numbers you substract** 
    * 37-(-27) = 64C
    * Difference in temperature is 64 degrees
    
# Chapter 7: Fractions

* Fractions represent part of something, part of a whole.
* Natural numbers, 1 2 3 4 5, are good for counting things. But what if something is less than 1? Then we use fractions! 
* Fractions are good for **measuring** and **comparing things**, while integers are good for counting things. 
    * Fractions are especially handy in comparing two different total amounts, i.e. if you want to compare the success of schools with a different amount of pupils. 
* Fractions break something up into **equal parts**, like an apple into half (1/2th)

## Introduction

* **numerator** = number of parts/slices/pieces
* **denominator** = number of equal parts/pieces/slices the whole is broken/sliced up into

### Whole Numbers as Fractions

* A pizza is split into 3 parts. You eat all 3 parts. How much have you eaten? 
    * 1 whole
    * Three thirds
    * 3/3
    
* **Fractions on a number line**
    * **3/5** = less than one on a number line
    * **1/1** = equal to one on the number line
    * **3/1** = more than one on a number line, it is 3! 
        * Notice the numerator is now larger than the denominator 
        * 1/1 + 1/1 + 1/1 = 3/1 
    * 4/4 = 1 whole, 1 on a number line
    * 4/1 = 4 wholes, 4 on a numbe rline


## Recognizing Fractions

* **What Fraction does the shaded area represent?**
    1. See how many pieces are shaded in 
    2. Add those pieces together = numerator
    3 denominator = the amount of pieces the whole is divded into, even if there is more than one whole shaded in
    
## Fractions on the number line

* Fractions are located between 0 and 1 on the number line
    * Unless they're bigger than 1
* Denomiator: Fractions split the space between 0 and 1 into equal pieces
    * 1/2 = splits the number 1 into 2 equal pieces
    * 1/5 = splits the number 1 into 5 equal pieces
    * **The size of each step along the number OR the number of steps between 0 and 1 = denomiator**
* Numerator: The posistion on the number line of the fraction is indicated by the numertor
    * 1/2 = the first split of the 2 splits
    * 1/5 = the first split of the 5 splits
    * 3/5 = the third split of the 5 splits
    * **The number of steps from zero = numerator**
* What if the numerator is bigger than the denominator?
    * The posistion on the number line is past one
    * The posistion carries on taking steps past 1. Each step is still the same length. 
    * In reality this means you have more than one or the whole. i.e. you have more than one apple pie, you have one apple pie and a extra slice! 
        * 1/4 = one split into 4 pieces, each located 0.25 places along.
        * 2/4 = 0.50. two slies of apple pie
        * 3/4 = 0.75. 3 slices of apple pie
        * 4/4 = 1. Whole Apple pie
        * 5/4 = 1.25. Whole apple pie and one slice. 
        * 6/4 = 1.50. Whole apple pie and 2 slices. 
        
* **examples of fractions on a number line**
    * 3/8 = The video game is split into 8 levels. There are 8 levels between 0 and 1. 1 = the whole game. You are 3 levels past zero.     
    * 5/4 = The apple pie is split into four slices. There are 4 pieces between 0 and 1. 1 = a whole pie. You have a whole apple pie and one extra slice. 
    
## Improper and Mixed Fractions

* **improper** = a fraction whose numerator is bigger than the denominator, i.e. a pie and one slice.
    * 3/2

* **mixed** = a whole number and a fraction, i.e. a whole pie and one slice
    * 1 1/2

* **mixed and improper fractions are just different ways of writing the same measurement, changing a fraction from mixed to improper does not change the value**

* **Converting an Improper to a Mixed Fraction**
    1. How many times does the denominator go into the numerator?
        * 5/4 = 1 r1 
    2. Add on the remainer as an an extra piece/slice
        * 5/4 = 1 1/4
    3. Both fractions show the same measurement, **a whole pie with one extra slice** or 5 slices in total (where 4 slices makes a pie)
* **Converting a Mixed fraction to a Improper**
    1. multiply the whole number by the denominator
        * 3 1/2 = 3*2/2
        * ***note***: another denominator goes under the whole number, making it into a fration
    2. Add on the numerator
        * 3 1/2 = (3*2)+1/2
        * 3 1/2 = 7/2
    3. Both fractions show the same measurement, **3 whole pieces and one extra slice** or 7 slices in total (where 2 slices makes a pie)

```{r fig.height = 3, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("grade6_fig/C7_F4.PNG"))
```
 
* **Why does this work?**
    * example 1: 5 1/4 
        * The 5 is the same thing as 20/4
        * This make the fraction $20/4 + 1/4 = 21/4$
        * The 5 can be thought as 5 wholes, each split into four pieces. Count the pieces and you get 20 pieces. 
    * example 2: 3 1/2 
        * The 3 is the same thing as 6/2
        * This make the fraction $6/2 + 1/2 = 7/2$
        * The 3 can be fought of as 3 wholes, each split into 2 pieces. Count the pieces and you get 6 pieces. 
    
```{r fig.height = 3, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("grade6_fig/C7_F2.PNG"))
```

## Equivalent Fractions

* **equivalent fraction** = fractions with the same value
* **On the number line** = fraction with the same location

```{r fig.height = 3, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("grade6_fig/C7_F5.PNG"))
```

* **Equivalent Fraction Models**
* Complete the equation 8/12 = ?/3
    * Model A(1/12) has **four** pieces, for each piece of model B(1/3)
    * divide Model A's numerator and denomator by four = 8/12 -> 2/3
    * 8/12 = 2/3
    
* Complete the equation. 9/12 = ?/8
    * Model A(1/12) has 1.5 pieces for each piece of model B(1/8)
        * 12/8 = 1.5
    * divide numerator and denom of model A by 1.5 = 6/8
    * 9/12 = 6/8

```{r fig.height = 3, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("grade6_fig/C7_F6.PNG"))
```


* **Identifying equivalent fractions**
    * **step one**: Are the denominators the same?
        * Yes: Compare Numerators 
        * No: Can you divide the denominator of one model to match the other? I.e. can you get the denominators the same. 
            * i.e. 4/6 and 2/3. Divide the numerator of 4/6 by 2 to match 2/3.
            * yes: Do the fractions match?
            * no: probably not equivalent 
